PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-08-05 main.tex
More edits through sections 1-3

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergoes a large-scale instability
\citep{chambers98}.
%TRQ "Suddenly" seems a drastic term to use here.
Suddenly, the oligarchics are no longer on isolated, stable orbits and
coalesce to form Earth-sized planets through a series of extremely
energetic, giant impacts \citep{kokubo02, raymond05, raymond06}.
%TRQ new paragraph here?
Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received the most attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, falls into an awkward in-between, where there are too
many particles to directly track with traditional N-body codes, while
the gravitational influence of the few, but massive oligarchs that
form make statistical methods an inappropriate approach. Because of
this limitation, planetesimal accretion is usually modeled in a narrow
ring, and the results are then scaled to suit whatever situation is being studied. N-body simulations of terrestrial planet formation typically begin with a series of neatly spaced oligarchs, whose mass varies smoothly with orbital period. As we will show in this paper, this extrapolation with orbital period is inappropriate, particularly at short ($<$ 10 day) periods.

@@

We begin our analysis by considering a disk of equal planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$, $\Gamma$ is the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
% TRQ: mention that $v$ is related to the velocity dispersion of the
% planetesimals.  Also: mention that the scale height of the disk is
% assumed to be \Omega/(2v)
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}

In the case where the collision cross section, $\Gamma$, depends only
on the physical size of the planetesimals, the growth scales linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \Gamma_{geo} \left( 1 + v_{esc}^2 / v^2 \right)$, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.
%TRQ: reference Safranov?

@@

An important feature is missing from the model described above, which
limits its applicability at late times. Encounters between
planetesimals that do not result in a collision play a crucial role
over long timescales. These close encounters act to convert energy
from Keplerian shear into random motion. Over time, this gravitational
stirring effect will raise the typical encounter velocities between
bodies and diminish the effectiveness of gravitational focusing. With
a spectrum of masses, these velocity differences become even more
pronounced because as the system evolves, it moves to a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
%TRQ: changed this to match Ida & Makino 1993.
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 10$ for a planetesimal disk. Despite
the fact that the behavior of gravitational stirring is well-described
by the two-body formalism, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the main pathway for energy exchange between planetesimals is unable to operate.

As mentioned above, the system will tend toward a state of energy
equipartition for a non-uniform mass distribution. \citet{kokubo98}
showed that runaway growth is actually self-limiting. As the runaway
bodies grow larger, they become more effective at heating the
remaining planetesimals, which diminishes the effectiveness of
gravitational focusing and throttles the growth rate. Around the time
that the massive runaway bodies make up about 50 percent of the solid
surface density
%TRQ: citation needed.
(citation), a period of less vigorous ``oligarchic'' growth commences, in which the few largest bodies continue to accrete planetesimals at similar rates.

The picture described above relies upon an important factor, which is that the mass distribution evolves slow enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial planet forming region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by
$$
v = \left( 5/8 \left< e^2 \right>^{1/2} + \left< i^2 \right>^{1/2}
\right)^{1/2} v_{k},
$$
(citation) where $\left< e^2 \right>^{1/2}$ and $\left< i^2
\right>^{1/2}$ are the eccentricity and inclination dispersions of the
planetesimal disk and $v_{k}$ is the Keplerian velocity. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is described
in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

These scaling considerations motivate the range of parameters in the
following numerical experiments as explore the different regimes.

% Need a transition to next section here.

% analytic conditions for runaway and orderly growth
% not clear what happens as gravitational focusing factor approaches 1

% mass and velocity distributions both evolve, two-body relaxation time works to describe velocity evolution when dispersion dominated
% if relaxation time shorter than collision time, oligarchic growth operates
% but what happens if these timescales are flipped? not clear, because the timescales themselves depend on mass and velocity, which both evolve with time

% a cleaner formalism: compare planetesimal size to hill and gravitational radius (alpha, beta)
% alpha does not evolve with time, beta will tend towards 1

% motivation for n-body simulations: what do masses and orbital properties of embryos look like as we change alpha and beta?
% at short period, alpha and beta are large

@@

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \citet{richardson94} and \citet{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius XXX
$fill in later$ are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two children. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\citep{wallace19}.
Particles are advanced in time on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $5 \times 10^{23}$ g are placed with semimajor
axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. The argument of perihelion $\omega$, longitude of ascending node
$\Omega$, and mean anomaly M for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclination dispersion $\left<
i^{2} \right>^{1/2}$ is chosen to be half the eccentricity
dispersion $\left< e^{2} \right>^{1/2}$ \citep{ida93a}.
% Mention Rayleigh distribution: TRQ: yes!

@@

\begin{table}
% I'm going to replace the fdLo simulation with a lower resolution version m_pl = 5e22 for consistency
% Originally wanted to run everything at m_pl = 1e22, which corresponds to 100km bodies, but this wasn't feasible at higher Sigma
\caption{Summary of Full Disk Simulations Run}
\begin{tabular}{llllll} \hline \hline
Name    & $m_{pl}^{a}$ & $N_{pl}^{b}$ & $A^{c}$ & $\alpha^{d}$ &  \\ \hline
fdHi                 & $5 \times 10^{22}$ & 903,687 & 100                                                         & 1.5                              &   \\
fdHiShallow    & $5 \times 10^{22}$ & 903,687 & 100                                                        & 0.5                              &   \\
fdHiSteep       & $5 \times 10^{22}$ & 903,687 & 100                                                         & 2.5                              &   \\
fdLo                & $1 \times 10^{22}$ & 45,185    & 1                                                             & 1.5                              &   \\ \hline
\end{tabular}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Planetesimal mass [g] \\
		       $^b$ Number of planetesimals \\
		       $^c$ Solid surface density normalization (relative to MMSN) \\
		       $^d$ Solid surface density power law index } \\
\end{flushleft}
\label{tab:sims}
\end{table}

>>>>

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergoes a large-scale instability
\citep{chambers98}.
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets'
through a series of extremely energetic, giant 
impacts \citep{kokubo02, raymond05, raymond06}.

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received the most attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, falls into an awkward in-between, where there are too
many particles to directly track with traditional N-body codes, while
the gravitational influence of the few, but massive oligarchs that
form make statistical methods an inappropriate approach. Because of
this limitation, planetesimal accretion is usually modeled in a narrow
ring, and the results are then scaled to suit whatever situation is being studied. N-body simulations of terrestrial planet formation typically begin with a series of neatly spaced oligarchs, whose mass varies smoothly with orbital period. As we will show in this paper, this extrapolation with orbital period is inappropriate, particularly at short ($<$ 10 day) periods.

@@

We begin our analysis by considering a disk of equal planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $\Omega/(2v)$. $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to the rms velocity, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}

In the case where the collision cross section, depends only
on the physical size of the planetesimals, the growth scales linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \Gamma_{geo} \left( 1 + v_{esc}^2 / v^2 \right)$ \citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

@@

An important feature is missing from the model described above, which
limits its applicability at late times. Encounters between
planetesimals that do not result in a collision play a crucial role
over long timescales. These close encounters act to convert energy
from Keplerian shear into random motion. Over time, this gravitational
stirring effect will raise the typical encounter velocities between
bodies and diminish the effectiveness of gravitational focusing. With
a spectrum of masses, these velocity differences become even more
pronounced because as the system evolves, it moves to a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 10$ for a planetesimal disk. Despite
the fact that the behavior of gravitational stirring is well-described
by the two-body formalism, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the main pathway for energy exchange between planetesimals is unable to operate.

As mentioned above, the system will tend toward a state of energy
equipartition for a non-uniform mass distribution. \citet{kokubo98}
showed that runaway growth is actually self-limiting. As the runaway
bodies grow larger, they become more effective at heating the
remaining planetesimals, which diminishes the effectiveness of
gravitational focusing and throttles the growth rate. Around the time
that the mass of the runaway bodies exceeds the mass of the planetesimals
by a factor of $\sim 50-100$ (depending on the orbital period)
\citep{ida93} a phase of less vigorous ``oligarchic'' 
growth commences, in which the few largest bodies continue to 
accrete planetesimals at similar rates.

The picture described above relies upon an important assumption, which is that the mass distribution evolves slow enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial planet forming region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by
$$
v = \left( 5/8 \left< e^2 \right>^{1/2} + \left< i^2 \right>^{1/2}
\right)^{1/2} v_{k},
$$
\citep{lissauer93} where $\left< e^2 \right>^{1/2}$ and $\left< i^2
\right>^{1/2}$ are the eccentricity and inclination dispersions of the
planetesimal disk and $v_{k}$ is the Keplerian velocity. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is described
in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

These scaling considerations motivate the range of parameters in the
following numerical experiments as we explore the different regimes.

@@

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \citet{richardson94} and \citet{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius
$2 \Delta T v + 2 r_{pl}$, where $\Delta T$ is the current timestep size
for the particle and $v$ is magnitude of its heliocentric velocity,
 are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two children. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\citep{wallace19}.
Particles are advanced in time on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $5 \times 10^{23}$ g are placed with semimajor
axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. The argument of perihelion $\omega$, longitude of ascending node
$\Omega$, and mean anomaly M for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclinations and eccentricities are drawn
from a Rayleigh distribution with
$\left< i^{2} \right> = 1/2 \left< e^{2} \right>$ \citep{ida93a}.

@@

\begin{table}
% I'm going to replace the fdLo simulation with a lower resolution version m_pl = 5e22 for consistency
% Originally wanted to run everything at m_pl = 1e22, which corresponds to 100km bodies, but this wasn't feasible at higher Sigma
\caption{Summary of Full Disk Simulations Run}
\begin{tabular}{llllll} \hline \hline
Name    & $m_{pl}^{a}$ & $N_{pl}^{b}$ & $A^{c}$ & $\alpha^{d}$ &  \\ \hline
fdHi                 & $5 \times 10^{22}$ & 903,687 & 100                                                         & 1.5                              &   \\
fdHiShallow    & $5 \times 10^{22}$ & 903,687 & 100                                                        & 0.5                              &   \\
fdHiSteep       & $5 \times 10^{22}$ & 903,687 & 100                                                         & 2.5                              &   \\
fdLo                & $5 \times 10^{22}$ & 45,185    & 1                                                             & 1.5                              &   \\ \hline
\end{tabular}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Planetesimal mass [g] \\
		       $^b$ Number of planetesimals \\
		       $^c$ Solid surface density normalization (relative to MMSN) \\
		       $^d$ Solid surface density power law index } \\
\end{flushleft}
\label{tab:sims}
\end{table}
